---
title: APSTA-GE 2011. Homework \#2
author: "January 2025: Yael Beshaw"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, fig.width = 4, 
                      fig.height = 4, fig.align = "center")
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r libs,include=TRUE, message= FALSE}
#likely useful libraries and initial seed set.
library(caret)
library(cluster)
library(klaR)
library(ggplot2)
library(ggdendro)
library(GGally)
library(e1071)
library(knitr)
library(foreign)
library(fpc)
library(gridExtra)
library(palmerpenguins)
set.seed(2011)
```

This assignment uses the \texttt{penguins} dataset.  The written work to hand in
consists of selected output from the software package, which you should include 
in a PDF document as you answer the questions.

```{r, include=TRUE}
data(penguins)
penguins <- penguins[complete.cases(penguins), ]

#standardize the data
penguins.stdz <- penguins;penguins.stdz[,3:5] <- scale(penguins[,3:5])
```

You will need to understand k-means clustering and choosing the number of 
clusters using the Calinski-Harabasz, or C(g) criterion, discussed in Handout 
2 and in class.

# **Q1**: 

Use k-means clustering (algorithm=Hartigan-Wong), finding solutions 
for g=2,3,4,5,6,7,8,9 groups.  ALWAYS start each clustering run by 
set.seed(2011) before each call to kmeans AND add this option to the call: 
nstart=100 (this generates 100 random starts and keeps kmeans from producing 
local minimum solutions). \textcolor{red}{Remember: you do not need to include 
the body mass measure.} [0 pts, but you must do it to proceed]

```{r, include=TRUE}
#specify the columns of interest
penguins.stdz_data <- penguins.stdz[, c("bill_length_mm", "bill_depth_mm", 
                           "flipper_length_mm")]


#loop through the solutions for g and store the results
kmeans_g <- list()

for (g in 2:9) {
  set.seed(2011)
  penguinz_kmeans <- kmeans(penguins.stdz_data,
                          centers= g,
                          nstart= 100,
                          algorithm = "Hartigan-Wong")
  kmeans_g[[(g)]] <- penguinz_kmeans
  
}
```

# **Q2**:	

For each value of g, compute $C(g)=(\sum \mbox{msb})/(\sum \mbox{msw})$.
In R, use the \texttt{calinhara} function given in package \texttt{fpc}.  For
this assignment, do not use the \texttt{NbClust} function in 
the \texttt{NbClust} R library, because it runs the kmeans fits for you, 
which will make it harder to generate them equivalently afterward. 
PLOT $C(g)$ as a function of $g$. [2pts]

```{r, include=TRUE}

# create a list that stores the clusters for each g that we will
# run the calinhara function on

Cg_kmeans <- list()

for (g in 2:9) {
  kmeans <- kmeans_g[[(g)]]$cluster
  Cg_kmeans[[g]] <- kmeans
}


# store the results of the Cg for each in a list
Cg_g <- list()

# run the calinhara() function for each g 
for (g in 2:9) {
  set.seed(2011)
  Cg <- calinhara(penguins.stdz_data,
                  Cg_kmeans[[g]],
                  cn=max(Cg_kmeans[[g]]))
  Cg_g[[(g)]] <- Cg
}

# turn into table in order to plot C(g) as a function of g
Cg <- data.frame(
  `g` = 2:9,
  `Cg` =  unlist(Cg_g)
)

#plot 
ggplot(Cg, aes(x = g, y = Cg)) +
  geom_line() +
  labs(title = "Question 2: Plot of C(g) vs Clusters (g)",
       x= "Number of Clusters",
       y = "C(g)") +
  theme_minimal()

```

# **Q3**:	

Choose the $g$ such that $C(g)$ is maximized.  Call this 
variable $g_{opt}$.  [1pt]
```{r, include=TRUE}
# The g such that C(g) is maximized is 3
g_opt <- 3
```

g = 3

# **Q4**: 

Compute the k-means cluster solutions using the following number of 
groups: $g_{opt}$, $g_{opt}+1$, and $g_{opt}+2$; compare the results using 
crosstabs (ignoring the $g_{opt}+1$ to $g_{opt}+2$ comparison) 
[1 pt each; 2 pts total] and comment on the effect of changing the number of 
clusters for each crosstab. [0.5pts each; 1pt total]

```{r, include=TRUE}
# compute the kmeans cluster solutions
km.penguins.opt <- kmeans(penguins.stdz_data, g_opt)
km.penguins.opt1 <- kmeans(penguins.stdz_data, g_opt + 1)
km.penguins.opt2 <- kmeans(penguins.stdz_data, g_opt + 2)

# compare the results using crosstabs
xtabs(~ km.penguins.opt$cluster + km.penguins.opt1$cluster)
xtabs(~ km.penguins.opt$cluster + km.penguins.opt2$cluster)
```
When analyzing the effect of adding more clusters, we see that in the 
first cluster from our g_opt = 3 is defined in both g_opt + 1 and in
g_opt + 2 with 119 penguins. However, in the comparison of g_opt and 
g_opt + 1, there is a split occurring with g_opt's second cluster in which
g_opt + 1 separates the 146 penguins into two separate clusters almost
down the middle. It also separates g_opt's third cluster into 2 clusters
but with only 6% of the penguins in a separate cluster. Thus, we have
evidence that g_opt's Cluster 2 could be divided further. When we assess
the comparision of g_opt and g_opt + 2, we see a similar story. The second
cluster of g_opt is again split across 3 different clusters with near
equal separation whereas Cluster 3 is not seperated much. This
supports the conclusion that while 3 clusters is the most optimal according
to C(g), there is evidence that we may have a basis for further
dividing Cluster 2, resulting in 4 clusters instead.

# **Q5** 

Color the bivariate plot of bill depth (y axis) vs. bill length (x-axis)
for each of the three solutions with $g=\{g_{opt},g_{opt}+1,g_{opt}+2\}$ and 
describe what changes with each solution. 
[1 pt ea. plot; 1 pt ea. comment; 6 pts total]

## g = g_opt

```{r, include=TRUE}
# g = g_opt
km.penguins.opt <- km.penguins.opt$cluster
penguins.stdz_data$km.penguins.opt <- km.penguins.opt
ggplot(data = penguins.stdz_data, aes(x=bill_length_mm, y= bill_depth_mm, 
                              color= factor(km.penguins.opt))) +
  geom_point() +
  labs(
    title = 
      "Q5: Bill Length vs Depth for g_opt",
    x= "Bill Length",
    y= "Bill Depth",
    color = "Cluster"
  )
```
Taking a look here at the bivariate plot for Bill Depth vs Length, we
can see that the three clusters are defined with overlap in between which
is expected. Cluster 2 (green) seems to have the most observations with 
minimal overlap into Clusters 1 or 3. Overall, 3 clusters provides the most
distinct cluster groups and is optimal based on our definition of C(g).


## g = g_opt + 1

```{r}
# g = g_opt + 1
km.penguins.opt1 <- km.penguins.opt1$cluster
penguins.stdz_data$km.penguins.opt1 <- km.penguins.opt1
ggplot(data = penguins.stdz_data, aes(x=bill_length_mm, y= bill_depth_mm, 
                              color= factor(km.penguins.opt1))) +
  geom_point() +
  labs(
    title = 
      "Q5: Bill Length vs Depth for g_opt + 1",
    x= "Bill Length",
    y= "Bill Depth",
    color = "Cluster"
  )
```
Once we add another cluster to our data, we see that Cluster 2 in g_opt is now
split into 2 separate clusters (1 and 4). We see here that this split is almost
right down the middle and provides evidence that while these groups are similar,
they have a distinctive feature that we can investigate to assess whether or not
4 clusters is more in line with our data and the questions we are trying to 
answer.


## g = g_opt + 2

```{r}
# g = g_opt + 1
km.penguins.opt2 <- km.penguins.opt2$cluster
penguins.stdz_data$km.penguins.opt2 <- km.penguins.opt2
ggplot(data = penguins.stdz_data, aes(x=bill_length_mm, y= bill_depth_mm, 
                              color= factor(km.penguins.opt2))) +
  geom_point() +
  labs(
    title = 
      "Q5: Bill Length vs Depth for g_opt + 2",
    x= "Bill Length",
    y= "Bill Depth",
    color = "Cluster"
  )
```

Finally, once we split the data into 5 clusters. Our original Cluster 1 and 3 
remain unchanged for the exception of 2-4 observations each. More importantly, 
we see that our orginal Cluster 2 is now split into three clusters. This 
clustering is not as clear and informative as g_opt or g_opt + 1, but it gives
us insight about the original Cluster 2 possibly being oversimplified. 
